Drought regimens predict life history strategies in \emph{Heliophila}

### 2019-06-22 manuscript/manuscript.tex
minor spelling

To characterize the distributions of annual (studied here, n = 21) and
perennial (studied here, n = 21) \emph{Heliophila} species, all (8670)
records for the genus \emph{Heliophila} were downloaded from the Global
Biodiversity Information Facility (gbif.org) on July 21, 2018 (GBIF,
2018). Herbaria records such as these provide a rich data sources to
characterize the geographical distributions of species (Thiers, 2016;
Willis \emph{et al.}, 2017; Lang \emph{et al.}, 2018). And as they
become digitized (Soltis, 2017), herbaria collections have been used to
study relationships between trait distributions, geography, and climate
(Davis \emph{et al.}, 2015; Stropp \emph{et al.}, 2016; Wolf \emph{et
al.}, 2016; Václavı'k \emph{et al.}, 2017).

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{../figures/maps_boxplots} \caption{Locations of occurrence records of (a) annual and (b)
perennial \emph{Heliophila.} Drought frequency during the (c) winter,
(d) spring, (e) summer and (f) autumn detected using the VHI. (g)
Drought frequencies during each season at the observation locations of
annual and perennial \emph{Heliophila} (Post-hoc contrasts annuals and
perennials from ANOVA, ** = p \textless{} 0.01).}\label{fig:mapsboxplots}
\end{figure}

@@

\begin{table}[tbp]
\begin{center}
\begin{threeparttable}
\caption{\label{tab:modelstable}Phylogenetic logistic regressions between life history, and the mean drought frequency observed at specimen sites of Heliophila species the winter, spring, summer, and autumn}
\begin{tabular}{lll}
\toprule
Predictor & \multicolumn{1}{c}{Estimate} & \multicolumn{1}{c}{P}\\
\midrule
Intercept & 0.7231 & 0.6636\\
Winter drought freq. & -1.5452 & 0.7274\\ \midrule
Intercept & 5.0107 & 0.0534\\
Spring drought freq. & -12.9014 & 0.0464\\ \midrule
Intercept & 7.7093 & 0.0054\\
Summer drought freq. & -19.9056 & 0.0042\\ \midrule
Intercept & 7.0162 & 0.0082\\
Fall drought freq. & -20.8174 & 0.0067\\ \midrule
\bottomrule
\addlinespace
\end{tabular}
\begin{tablenotes}[para]
\normalsize{\textit{Note.} Annual species were scored as 0 and perennial species as 1.}
\end{tablenotes}
\end{threeparttable}
\end{center}
\end{table}

@@

These findings suggest that species with locally adaptive life history
strategies could be threatened by rapidly changing drought regimens
(Dai, 2011). In light of the findings here, forecasted reductions in
rainfall across eastern South Africa (Service \& Comission, 2017) could
be particularly impactful to plant community compositions. Here we found
that this region is currently dominated by derived perennial species of
\emph{Heliophila}. However, a scenario in which droughts become more
frequent in this region may allow for the establishment of annuals. Such
changes in selection patterns and shifts in plant functional diversity
could have impacts on ecosystem functioning and processes such as carbon
cycling (Garnier \emph{et al.}, 1997; Roumet \emph{et al.}, 2006; Monroe
\emph{et al.}, 2018a). Furthermore, the changes in frequency of drought
may be an important factor when considering the use of perennial
cropping systems (Parry \emph{et al.}, 2005; Lelièvre \& Volaire, 2009).

@@

\leavevmode\hypertarget{ref-clausen1948experimental}{}%
\textbf{\textnormal{Clausen J}, \textnormal{Keck DD}, \textnormal{Hiesey
WM}, \textnormal{others}}. \textbf{1948}. Experimental studies on the
nature of species. III. Environresponses of climatic races of achillea.
\emph{Experimental studies on the nature of species. III.
Environresponses of climatic races of Achillea.}

\leavevmode\hypertarget{ref-cody1975ecology}{}%
\textbf{\textnormal{Cody ML}, \textnormal{MacArthur RH},
\textnormal{Diamond JM}, \textnormal{others}}. \textbf{1975}.
\emph{Ecology and evolution of communities}. Harvard University Press.

@@

\leavevmode\hypertarget{ref-lang2018using}{}%
\textbf{\textnormal{Lang PL}, \textnormal{Willems FM},
\textnormal{Scheepens J}, \textnormal{Burbano HA}, \textnormal{Bossdorf
O}}. \textbf{2018}. \emph{Using herbaria to study global environmental
change}. PeerJ Preprints.

@@

\leavevmode\hypertarget{ref-Mojica2016}{}%
\textbf{\textnormal{Mojica JP}, \textnormal{Mullen J},
\textnormal{Lovell JT}, \textnormal{Monroe JG}, \textnormal{Paul JR},
\textnormal{Oakley CG}, \textnormal{McKay JK}}. \textbf{2016}. Genetics
of water use physiology in locally adapted Arabidopsis thaliana.
\emph{Plant Science}.

@@

\leavevmode\hypertarget{ref-monroe2018drought}{}%
\textbf{\textnormal{Monroe J}, \textnormal{Powell T}, \textnormal{Price
N}, \textnormal{Mullen J}, \textnormal{Howard A}, \textnormal{Evans K},
\textnormal{Lovell J}, \textnormal{McKay J}}. \textbf{2018b}. Drought
adaptation in nature by extensive genetic loss-of-function.
\emph{eLife}: DOI: 10.7554/eLife.41038.

@@

\leavevmode\hypertarget{ref-mummenhoff2005phylogeny}{}%
\textbf{\textnormal{Mummenhoff K}, \textnormal{Al-Shehbaz IA},
\textnormal{Bakker FT}, \textnormal{Linder HP}, \textnormal{Mühlhausen
A}}. \textbf{2005}. Phylogeny, morphological evolution, and speciation
of endemic brassicaceae genera in the cape flora of southern africa.
\emph{Annals of the Missouri Botanical Garden}: 400--424.

@@

\leavevmode\hypertarget{ref-raunkiaer1905types}{}%
\textbf{\textnormal{Raunkiaer CC}}. \textbf{1905}. \emph{Types
biologiques pour la géographie botanique}.

@@

\leavevmode\hypertarget{ref-sayre2013new}{}%
\textbf{\textnormal{Sayre RG}, \textnormal{Comer P}, \textnormal{Hak J},
\textnormal{Josse C}, \textnormal{Bow J}, \textnormal{Warner H},
\textnormal{Larwanou M}, \textnormal{Kelbessa E}, \textnormal{Bekele T},
\textnormal{Kehl H} \emph{et al.}} \textbf{2013}. A new map of
standardized terrestrial ecosystems of africa. \emph{African
Geographical Review}.

\leavevmode\hypertarget{ref-schaffer1975selection}{}%
\textbf{\textnormal{Schaffer W}, \textnormal{Gadgil M}}. \textbf{1975}.
Selection for optimal life histories in plants. \emph{Ecology and
evolution of communities.}: 142--157.

\leavevmode\hypertarget{ref-south2017a}{}%
\textbf{\textnormal{Service SAW}, \textnormal{Comission WR}}.
\textbf{2017}. \emph{A climate change reference atlas}.

@@

\leavevmode\hypertarget{ref-stearns1992evolution}{}%
\textbf{\textnormal{Stearns SC}}. \textbf{1992}. \emph{The evolution of
life histories}.

@@

\leavevmode\hypertarget{ref-willis2017old}{}%
\textbf{\textnormal{Willis CG}, \textnormal{Ellwood ER},
\textnormal{Primack RB}, \textnormal{Davis CC}, \textnormal{Pearson KD},
\textnormal{Gallinat AS}, \textnormal{Yost JM}, \textnormal{Nelson G},
\textnormal{Mazer SJ}, \textnormal{Rossington NL} \emph{et al.}}
\textbf{2017}. Old plants, new tricks: Phenological research using
herbarium specimens. \emph{Trends in ecology \& evolution} \textbf{32}:
531--546.

@@

\hypertarget{images-used}{%
\subsubsection{Images used}\label{images-used}}

\url{https://www.gbif.org/occurrence/1099023487}\\
\url{https://www.gbif.org/occurrence/1057389408}~\\
\url{https://www.gbif.org/occurrence/1099023562}~\\
\url{https://www.gbif.org/occurrence/1099023490}

>>>>

To characterize the distributions of annual (studied here, n = 21) and
perennial (studied here, n = 21) \emph{Heliophila} species, all (8670)
records for the genus \emph{Heliophila} were downloaded from the Global
Biodiversity Information Facility (gbif.org) on July 21, 2018 (GBIF,
2018). Herbaria records such as these provide a rich data sources to
characterize the geographical distributions of species (Thiers, 2016;
Willis \emph{et al.}, 2017; Lang \emph{et al.}, 2019). And as they
become digitized (Soltis, 2017), herbaria collections have been used to
study relationships between trait distributions, geography, and climate
(Davis \emph{et al.}, 2015; Stropp \emph{et al.}, 2016; Wolf \emph{et
al.}, 2016; Václavı'k \emph{et al.}, 2017).

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{../figures/maps_boxplots} \caption{Locations of occurrence records of (a) annual and (b)
perennial \emph{Heliophila.} Drought frequency during the (c) winter,
(d) spring, (e) summer and (f) autumn detected using the VHI. (g)
Drought frequencies during each season at the observation locations of
annual and perennial \emph{Heliophila} (Post-hoc contrasts annuals and
perennials from ANOVA, ** = p \textless{} 0.01). Boxplots indicate
median (horizontal line), first and third quartiles (upper and lower
limits of box), and 1.5 times the interquartile range (whiskers).
Individual observations are shown by crosshairs. Annual and perennial
species are indicated by data colored blue and oranage, respectively.}\label{fig:mapsboxplots}
\end{figure}

@@

\begin{table}[tbp]
\begin{center}
\begin{threeparttable}
\caption{\label{tab:modelstable}Phylogenetic logistic regressions between life history, and the mean drought frequency observed at specimen sites of Heliophila species the winter, spring, summer, and autumn}
\begin{tabular}{lll}
\toprule
Predictor & \multicolumn{1}{c}{Estimate} & \multicolumn{1}{c}{P}\\
\midrule
Intercept & 0.7231 & 0.6636\\
Winter drought freq. & -1.5452 & 0.7274\\ \midrule
Intercept & 5.0107 & 0.0534\\
Spring drought freq. & -12.9014 & 0.0464\\ \midrule
Intercept & 7.7093 & 0.0054\\
Summer drought freq. & -19.9056 & 0.0042\\ \midrule
Intercept & 7.0162 & 0.0082\\
Autumn drought freq. & -20.8174 & 0.0067\\ \midrule
\bottomrule
\addlinespace
\end{tabular}
\begin{tablenotes}[para]
\normalsize{\textit{Note.} Annual species were scored as 0 and perennial species as 1.}
\end{tablenotes}
\end{threeparttable}
\end{center}
\end{table}

@@

These findings suggest that species with locally adaptive life history
strategies could be threatened by rapidly changing drought regimens
(Dai, 2011). In light of the findings here, forecasted reductions in
rainfall across eastern South Africa (SAWS \& WRC, 2017) could be
particularly impactful to plant community compositions. Here we found
that this region is currently dominated by derived perennial species of
\emph{Heliophila}. However, a scenario in which droughts become more
frequent in this region may allow for the establishment of annuals. Such
changes in selection patterns and shifts in plant functional diversity
could have impacts on ecosystem functioning and processes such as carbon
cycling (Garnier \emph{et al.}, 1997; Roumet \emph{et al.}, 2006; Monroe
\emph{et al.}, 2018a). Furthermore, the changes in frequency of drought
may be an important factor when considering the use of perennial
cropping systems (Parry \emph{et al.}, 2005; Lelièvre \& Volaire, 2009).

@@

\leavevmode\hypertarget{ref-clausen1948experimental}{}%
\textbf{\textnormal{Clausen J}, \textnormal{Keck DD}, \textnormal{Hiesey
WM}}. \textbf{1948}. Experimental studies on the nature of species. III.
Environresponses of climatic races of achillea. \emph{Experimental
studies on the nature of species. III. Environresponses of climatic
races of Achillea.}

\leavevmode\hypertarget{ref-cody1975ecology}{}%
\textbf{\textnormal{Cody ML}, \textnormal{MacArthur RH},
\textnormal{Diamond JM}}. \textbf{1975}. \emph{Ecology and evolution of
communities}. Harvard University Press.

@@

\leavevmode\hypertarget{ref-lang2019using}{}%
\textbf{\textnormal{Lang PL}, \textnormal{Willems FM},
\textnormal{Scheepens J}, \textnormal{Burbano HA}, \textnormal{Bossdorf
O}}. \textbf{2019}. Using herbaria to study global environmental change.
\emph{New Phytologist} \textbf{221}: 110--122.

@@

\leavevmode\hypertarget{ref-mojica2016genetics}{}%
\textbf{\textnormal{Mojica JP}, \textnormal{Mullen J},
\textnormal{Lovell JT}, \textnormal{Monroe JG}, \textnormal{Paul JR},
\textnormal{Oakley CG}, \textnormal{McKay JK}}. \textbf{2016}. Genetics
of water use physiology in locally adapted arabidopsis thaliana.
\emph{Plant Science} \textbf{251}: 12--22.

@@

\leavevmode\hypertarget{ref-monroe2018drought}{}%
\textbf{\textnormal{Monroe J}, \textnormal{Powell T}, \textnormal{Price
N}, \textnormal{Mullen J}, \textnormal{Howard A}, \textnormal{Evans K},
\textnormal{Lovell J}, \textnormal{McKay J}}. \textbf{2018b}. Drought
adaptation in arabidopsis thaliana by extensive genetic
loss-of-function. \emph{eLife}: DOI: 10.7554/eLife.41038.

@@

\leavevmode\hypertarget{ref-mummenhoff2005phylogeny}{}%
\textbf{\textnormal{Mummenhoff K}, \textnormal{Al-Shehbaz IA},
\textnormal{Bakker FT}, \textnormal{Linder HP}, \textnormal{Mühlhausen
A}}. \textbf{2005}. Phylogeny, morphological evolution, and speciation
of endemic brassicaceae genera in the cape flora of southern africa.
\emph{Annals of the Missouri Botanical Garden} \textbf{92}: 400--424.

@@

\leavevmode\hypertarget{ref-raunkiaer1905types}{}%
\textbf{\textnormal{Raunkiaer CC}}. \textbf{1905}. Types biologiques
pour la géographie botanique. \emph{Oversigt over Det Kongelige Danske
Videnskabernes Selskabs Forhandlinger}: 347--438.

@@

\leavevmode\hypertarget{ref-south2017a}{}%
\textbf{\textnormal{SAWS}, \textnormal{WRC}}. \textbf{2017}. \emph{A
climate change reference atlas}.

\leavevmode\hypertarget{ref-sayre2013new}{}%
\textbf{\textnormal{Sayre RG}, \textnormal{Comer P}, \textnormal{Hak J},
\textnormal{Josse C}, \textnormal{Bow J}, \textnormal{Warner H},
\textnormal{Larwanou M}, \textnormal{Kelbessa E}, \textnormal{Bekele T},
\textnormal{Kehl H et a}}. \textbf{2013}. \emph{A new map of
standardized terrestrial ecosystems of africa}.

\leavevmode\hypertarget{ref-schaffer1975selection}{}%
\textbf{\textnormal{Schaffer W}, \textnormal{Gadgil M}}. \textbf{1975}.
Selection for optimal life histories in plants. In: Cody M, Diamond J,
eds. The ecology and evolution of communities. Cambridge, MA: Belknap,
142--157.

@@

\leavevmode\hypertarget{ref-stearns1992evolution}{}%
\textbf{\textnormal{Stearns SC}}. \textbf{1992}. \emph{The evolution of
life histories}. Oxford: Oxford University Press.

@@

\leavevmode\hypertarget{ref-willis2017old}{}%
\textbf{\textnormal{Willis CG}, \textnormal{Ellwood ER},
\textnormal{Primack RB}, \textnormal{Davis CC}, \textnormal{Pearson KD},
\textnormal{Gallinat AS}, \textnormal{Yost JM}, \textnormal{Nelson G},
\textnormal{Mazer SJ}, \textnormal{Rossington NL et a}}. \textbf{2017}.
Old plants, new tricks: Phenological research using herbarium specimens.
\emph{Trends in ecology \& evolution} \textbf{32}: 531--546.

@@

\hypertarget{notes-s1}{%
\subsubsection{Notes S1}\label{notes-s1}}

Images of herbarium specimens in Figure 1. Used under creative commons
licenses, permissions not required.
\url{https://www.gbif.org/occurrence/1099023487}\\
\url{https://www.gbif.org/occurrence/1057389408}~\\
\url{https://www.gbif.org/occurrence/1099023562}~\\
\url{https://www.gbif.org/occurrence/1099023490}
